---
title: "Homework 3"
author: "Yuki Joyama"
output: 
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fontspec", "amsmath", "amssymb", "unicode-math"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r}
# libraries
library(tidyverse)
library(ggplot2)

# setup plot theme
theme_set(
  theme_bw() +
    theme(legend.position = "top")
  )

###CODE FOR HW3#####
set.seed(124)
n <- 16
p_C <- 1/5
C <- rbinom(n,1,p_C)
theta0 <- 1/2
theta1 <- -1/5
p_A <- theta0+theta1*C
A <- rbinom(n,1,p_A)
beta0 <- 110
beta1 <- 20
beta2 <- 5
sigma_Y <- 1
mu_Y <- beta0+beta1*C+beta2*A
Y <- rnorm(n,mu_Y, sigma_Y)
```


# 1. Interpret parameters
$C$: Obesity ($p=1/5$ of the mice are obese at baseline)    
$A$: Exposure to light  
$Y$: Glucose outcome such that $Y\sim N(\mu_Y, \sigma)$, where $E(\mu_Y)= f(\beta_0+\beta_1*obesity+\beta_2*light)$, $\beta_0=110, \beta_1=20, \text{ and } \beta_2=-5$   
  
$p$: The probability of a mouse to be obese at baseline. Theoretically, 3.2 out of 16 mice are obese. In the simulation by the above R code, 3 out of 16 mice are set to be obese at baseline.  
$\theta_0$: The probability for a non-obese mouse ($C=0$) to be exposed to light. $\theta_0=1/2$ means that there's a 50% chance for a non-obese mouse to be exposed to light.   
$\theta_1$: Describes how more (or less) likely it is for an obese mouse ($C=1$) to be exposed to light. $\theta_1=-1/5$ indicates that obese mice are 20% less likely to be exposed to the light. In other words, there's a 30% chance for an obese mouse to be exposed to light.     
$\beta_0$: The baseline mean glucose level for non-obese mice that are not exposed to light, which was set to be 110 mg/dL.   
$\beta_1$: The coefficient of obesity on glucose level. $\beta_1=20$ indicates that obese mice have 20 mg/dL higher average glucose level from the baseline compared to non-obese mice, holding other variables constant.       
$\beta_2$: The coefficient of light on glucose level. $\beta_2=-5$ suggests that mice with exposure to light have 5 mg/dL lower average glucose level from the baseline compared to non-exposed mice, holding other variables constant. 


# 2. PACE

# 3. g-formula (randomized vs. observational study)

# 4. Estimate and confidence interval of $E[Y|A=1]-E[Y|A=0]$

# 5. Estimate and confidence interval of $E[Y_1]-E[Y_0]$

# 6. Assumptions of estimate $E[Y_1]−E[Y_0]$ using linear regression

