---
title: "Homework 4"
author: "Yuki Joyama"
output: 
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fontspec", "amsmath", "amssymb", "unicode-math"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r}
# libraries
library(tidyverse)
library(ggplot2)
library(dagitty)
library(tableone)
library(knitr)
library(personalized)

# setup plot theme
theme_set(
  theme_minimal() +
    theme(legend.position = "top")
  )
```


```{r}
# import data 
df = read_csv("./data/hw4_data.csv") |> 
  mutate(
    treat = as.factor(treat),
    black = as.factor(black),
    hispan = as.factor(hispan),
    married = as.factor(married),
    nodegree = as.factor(nodegree)
  )
```

1. DAG
```{r}
g = dagitty('dag {
age [pos="-1.842,-0.039"]
black [pos="1.458,-0.317"]
educ [pos="-0.698,-0.114"]
hispan [pos="0.269,-0.760"]
married [pos="-2.461,-0.426"]
nodegree [pos="-1.567,-0.659"]
re74 [pos="-1.219,0.594"]
re75 [pos="0.011,0.971"]
re78 [outcome,pos="0.893,0.468"]
treat [exposure,pos="-2.558,0.500"]
age -> educ
age -> married
age -> re74
age -> re75
age -> re78
age -> treat
black -> educ
black -> nodegree
black -> re74
black -> re75
black -> re78
black -> treat
educ -> re74
educ -> re75
educ -> re78
educ -> treat
hispan -> educ
hispan -> nodegree
hispan -> re74
hispan -> re75
hispan -> re78
hispan -> treat
married -> treat
nodegree -> educ
nodegree -> re74
nodegree -> re75
nodegree -> re78
nodegree -> treat
re74 -> re75
re74 -> re78
re75 -> re78
treat -> re74
treat -> re75
treat -> re78
}')
plot(g)
```

Note of the variables in the DAG:  
`treat`: treatment assignment (job training); exposure  
- Suppose that the job training was completed before 1974, `treat` is likely to infludence `re74`, `re75` and `re78`  
`age`: age in years  
- `age` may affect `married`, `treat`, `educ` and all the income status (`re74`, `re75`, `re78`)    
`educ`: education in years   
- `educ` may affect all the income status and `treat`  
`black`, `hispan`: indicators for African American and hispanic  
- Both of the ethnicity indicator may affect all the income status, `nodegree`, `treat`, and `educ`  
`married`: indicator for married  
- `married` could influence `treat`  
`nodegree`: indicator for highschool degree  
- `nodegree` may affect `treat`, `educ`, and all the income status  
`re74`: income in 1974    
- `re74` can influence `re75` and `re78`  
`re75`: income in 1975  
- `re75` can influence `re78`  
`re78`: income in 1978; outcome    

Given the DAG, covariates that need to be adjusted in investigating the effect of exposure on the outcome are `nodegree`, `hispan`, `black`, `educ` and `age`.  

2. I will evaluate the covariate balance using standardized mean differences. 
```{r}
cov = c("nodegree", "hispan", "black", "educ", "age")

# construct a table
tab = CreateTableOne(vars = cov, strata = "treat", data = df, test = FALSE)
print(tab, smd = TRUE)
```
We can see that most covariates have SMD > 0.1 except for `educ`, indicating the potential imbalance between the two treatment group. 

3. Propensity score estimates are calculated by fitting a logistic regression. 
```{r}
# fit PS model
ps.fit <- glm(treat ~ nodegree + hispan + black + educ + age, family = "binomial", data = df)

ps.fit |> 
  broom::tidy() |> 
  kable()

# estimate PS
df.ps <- predict(ps.fit, type = 'response')
print(df.ps[1:50])
```
Listed values are the propensity scores for each observation in the dataset (only showing the first 50 observations out of 614).

4. 
```{r}
# append ps estimates to dataset
df <- df |> 
  mutate(
    ps.est = df.ps
  )

prop.func <- function(x, trt)
{
  # propensity score model
  propens.model <- glm(as.factor(treat) ~ as.factor(nodegree) + as.factor(hispan) + as.factor(black) + educ + age, family = "binomial", data = x)
  pi.x <- predict(propens.model, type = "response")
  pi.x
}

# histogram
check.overlap(x = df,
              trt = df$treat,
              propensity.func = prop.func)


# add density plot with histogram
check.overlap(x = df,
              trt = df$treat,
              type = "both",
              propensity.func = prop.func)
```
The histogram shows that the propensity score estimates does not overlap in ranges approximately [0, 0.0375] and [0.7375, 1]. I will trim the samples whose PS estimate is in these ranges. 

```{r}
# trimming
# eliminate controls for whom the P(A=1|C) is less that the min(P(A=1|C)) found in the treated group
min(df$ps.est[df$treat == 1])
head(df$ps.est[which(df$treat == 0)] <= min(df$ps.est[df$treat == 1]))

length(df$ps.est[which(df$treat == 0)]<= min(df$ps.est[df$treat == 1]))

# eliminate treated for whom the P(A=1|C) is greater that the max(P(A=1|C)) found in the control group
max(df$ps.est[df$treat == 0])
head(df$ps.est[which(df$treat == 1)]>= max(df$ps.est[df$treat == 0]))

length(df$ps.est[which(df$treat == 1)]>= max(df$ps.est[df$treat == 0]))

df_trim = df[df$ps.est >= min(df$ps.est[df$treat == 1]) & df$ps.est <= max(df$ps.est[df$treat == 0]),]

dim(df_trim)
```

After the trimming, sample size went down from 614 to 571, which may lead to reduced statistical power. In terms of the generalizability, trimming will disable the external validity because we are only looking at individuals with non-extreme propensity scores. 

5. Here I will reevaluate the covariate balance using the trimmed sample.
```{r}
# construct a table
tab_trim = CreateTableOne(vars = cov, strata = "treat", data = df_trim, test = FALSE)
print(tab_trim, smd = TRUE)
```
Compared to untrimmed data, the balance of `nodegree`, `black` and `age` appear to be improved while `educ` and `hispan` appear to be worsened.

6. Subclassification  
First I will refit the model using the trimmed data and recheck the overlap.
```{r}
# fit PS model
ps.fit_trim <- glm(treat ~ nodegree + hispan + black + educ + age, family = "binomial", data = df_trim)

ps.fit_trim |> 
  broom::tidy() |> 
  kable()

# estimate PS
df.ps_trim <- predict(ps.fit_trim, type = 'response')
print(df.ps_trim[1:50])
```

I will create three strata. The breaks are shown below.
```{r}
df_trim = df_trim |> 
  mutate(ps.strata = gtools::quantcut(ps.est, 3))

# breaks
quantile(df_trim$ps.est, probs = seq(0, 1, length.out = 4))
```

```{r}
breaks = quantile(df_trim$ps.est, probs = seq(0, 1, length.out = 4))

# save dataset for each strata
df_trim1 = df_trim |> 
  filter(ps.strata == "[0.0453,0.0759]")

df_trim2 = df_trim |> 
  filter(ps.strata == "(0.0759,0.585]")

df_trim3 = df_trim |> 
  filter(ps.strata == "(0.585,0.743]")
```

Below is the histogram and density plot of the propensity scores in each stratum. 
```{r}
# plot propensity scores with subclass boundaries by treatment group
# [0.0453,0.0759]
ggplot(df_trim1, aes(x = ps.est, fill = as.factor(treat))) +
  geom_histogram(aes(y = ..density..), bins = 30, alpha = 0.5, position = "identity") +
  geom_density(aes(color = as.factor(treat)), alpha = 0.5, size = 1) +  # add density plot
  scale_fill_manual(values = c("#F8766D", "#00BFC4"), name = "Treatment Group", labels = c("Control", "Treated"),
                    guide = guide_legend(override.aes = list(color = NA))) + 
  scale_color_manual(values = c("#F8766D", "#00BFC4"), guide = "none") +  
  geom_rug(aes(color = as.factor(treat)), sides = "b", alpha = 0.5) +  # add rug plot
  labs(title = "Distribution of Propensity Scores by Stratum 1",
       x = "Propensity Score",
       y = "Density") 

# (0.0759,0.585]
ggplot(df_trim2, aes(x = ps.est, fill = as.factor(treat))) +
  geom_histogram(aes(y = ..density..), bins = 30, alpha = 0.5, position = "identity") +
  geom_density(aes(color = as.factor(treat)), alpha = 0.5, size = 1) +  # add density plot
  scale_fill_manual(values = c("#F8766D", "#00BFC4"), name = "Treatment Group", labels = c("Control", "Treated"),
                    guide = guide_legend(override.aes = list(color = NA))) +  
  scale_color_manual(values = c("#F8766D", "#00BFC4"), guide = "none") +  
  geom_rug(aes(color = as.factor(treat)), sides = "b", alpha = 0.5) +  # add rug plot
  labs(title = "Distribution of Propensity Scores by Stratum 2",
       x = "Propensity Score",
       y = "Density")

# (0.585,0.743]
ggplot(df_trim3, aes(x = ps.est, fill = as.factor(treat))) +
  geom_histogram(aes(y = ..density..), bins = 30, alpha = 0.5, position = "identity") +
  geom_density(aes(color = as.factor(treat)), alpha = 0.5, size = 1) +  # add density plot
  scale_fill_manual(values = c("#F8766D", "#00BFC4"), name = "Treatment Group", labels = c("Control", "Treated"),
                    guide = guide_legend(override.aes = list(color = NA))) +  
  scale_color_manual(values = c("#F8766D", "#00BFC4"), guide = "none") +  
  geom_rug(aes(color = as.factor(treat)), sides = "b", alpha = 0.5) +  # add rug plot
  labs(title = "Distribution of Propensity Scores by Stratum 3",
       x = "Propensity Score",
       y = "Density") 
```
Now, let's inspect the covariate balance in each stratum. 

```{r}
# stratum 1
tab_trim1 = CreateTableOne(vars = cov, strata = "treat", data = df_trim1, test = FALSE)
print(tab_trim1, smd = TRUE)

# stratum 2
tab_trim2 = CreateTableOne(vars = cov, strata = "treat", data = df_trim2, test = FALSE)
print(tab_trim2, smd = TRUE)

# stratum 3
tab_trim3 = CreateTableOne(vars = cov, strata = "treat", data = df_trim3, test = FALSE)
print(tab_trim3, smd = TRUE)
```
Stratum 1 and 3 show acceptable balance for most covariates with `nodegree` being the most imbalanced one. Stratum 2 still shows significant imbalance, especially for the covariate `black` and `nodegree`. This may be due to the limited propensity score distribution within the range of (0.0759, 0.585].

7. 

8. 

9. The advantage of regression based approach is that it can be applied to 
