---
title: "Mediation Lab - Regression approach for causal mediation"
date: "12/07/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r message=F,warning=F}

######################################################
##   Statistical Methods for Causal Inference       ## 
##   R Application                                  ##
##   Lung cancer example                            ##
######################################################


## load in data 
data_sens <- read.csv("sim_data_lung.csv")

```{r fig.width=8,fig.height=4,message=F,warning=F}

######################################
### QUESTION 1C 2 AND 4###############
######################################

#install.packages("CMAverse")
library(CMAverse)

## run mediation analysis with interaction using cmest()
mediation.int.m1 <- cmest(data = data_sens, model = "rb", casecontrol = TRUE, yrare = TRUE,
                         outcome = "case", 
                         exposure = "snp", mediator = "smoking", EMint = TRUE,
                         prec = c("sex", "colgrad", "age"), 
                         mreg = list("logistic"), yreg = "logistic", 
                         a = 1, astar = 0, mval = list(1),
                         estimation = "paramfunc", inference = "delta", full=FALSE)

mediation.int.m0 <- cmest(data = data_sens, model = "rb", casecontrol = TRUE, yrare = TRUE,
                         outcome = "case", 
                         exposure = "snp", mediator = "smoking", EMint = TRUE,
                         prec = c("sex", "colgrad", "age"), 
                         mreg = list("logistic"), yreg = "logistic", 
                         a = 1, astar = 0, mval = list(0),
                         estimation = "paramfunc", inference = "delta", full=FALSE)

summary(mediation.int.m1)

ggcmest(mediation.int.m1) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, vjust = 0.8))

summary(mediation.int.m0)

ggcmest(mediation.int.m0) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, vjust = 0.8))


## CDE(1) = 1.53
## CDE(0) = 1.07
## NIE = 1.02 
## NDE = 1.45 

```

